---
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

There are 3-4 packages you will need to install for today's practical: `install.packages(c("xgboost", "eegkit", "forecast", "tseries", "caret"))` apart from that everything else should already be available on your system.

If you are using a newer Mac you may have to also install [quartz](https://www.xquartz.org/) to have everything work (do this if you see errors about `X11` during install/execution).

I will endeavour to use explicit imports to make it clear where functions are coming from (functions without `library_name::` are part of base R or a function we've defined in this notebook).

```{r libraries, echo=FALSE}
# Using the same library we used earlier in the course for tabular data because we know it works!
library(xgboost)

# EEG manipulation library in R (although very limited compared to signal processing libraries available in other languages, matlab might actually still be a leader in this specific area)
library(eegkit)

# some time series functions (that we only skim the depths of)
library(forecast)
library(tseries)
library(caret)

# just tidyverse libraries that should already be installed
library(dplyr)
library(reshape2)
library(purrr)
library(ggplot2)
```

## EEG Eye Detection Data

One of the most common types of medical sensor data (and one that we talked about during the lecture) are Electroencephalograms (EEGs).\
These measure mesoscale electrical signals (measured in microvolts) within the brain, which are indicative of a region of neuronal activity.
Typically, EEGs involve an array of sensors (aka channels) placed on the scalp with a high degree of covariance between sensors.

As EEG data can be very large and unwieldy, we are going to use a relatively small/simple dataset today from [this paper](http://ehrai.com/su/pdf/aihls2013.pdf).

This dataset is a 117 second continuous EEG measurement collected from a single person with a device called a "Emotiv EEG Neuroheadset".
In combination with the EEG data collection, a camera was used to record whether person being recorded had their eyes open or closed.
This was eye status was then manually annotated onto the EEG data with `1` indicated the eyes being closed and `0` the eyes being open.
Measures microvoltages are listed in chronological order with the first measured value at the top of the dataframe.

Let's parse the data directly from the `h2o` library's (which we aren't actually using directly) test data S3 bucket:

```{r parse_data}
eeg_url <- "https://h2o-public-test-data.s3.amazonaws.com/smalldata/eeg/eeg_eyestate_splits.csv"
eeg_data <- read.csv(eeg_url)

# add timestamp
Fs <- 117 / nrow(eeg_data)
eeg_data <- transform(eeg_data, ds = seq(0, 116.99999, by = Fs), eyeDetection = as.factor(eyeDetection))
print(table(eeg_data$eyeDetection))

# split dataset into train, validate, test
eeg_train <- subset(eeg_data, split == 'train', select = -split)
print(table(eeg_train$eyeDetection))

eeg_validate <- subset(eeg_data, split == 'valid', select = -split)
eeg_test <- subset(eeg_data, split == 'test', select = -split)
```

**0** Knowing the `eeg_data` contains 117 seconds of data, inspect the `eeg_data` dataframe and the code above to and determine how many samples per second were taken?

```{r}
samples_per_second <- nrow(eeg_data) / 117
print(samples_per_second)
```

**1** How many EEG electrodes/sensors were used?\
14 electrodes/sensors

### Exploratory Data Analysis

Now that we have the dataset and some basic parameters let's begin with the ever important/relevant exploratory data analysis.

First we should check there is no missing data!

```{r check_na}
sum(is.na(eeg_data))
```

Great, now we can start generating some plots to look at this data within the time-domain.

First we use `reshape2::melt()` to transform the `eeg_data` dataset from a wide format to a long format expected by `ggplot2`.

Specifically, this converts from "wide" where each electrode has its own column, to a "long" format, where each observation has its own row.
This format is often more convenient for data analysis and visualization, especially when dealing with repeated measurements or time-series data.

We then use `ggplot2` to create a line plot of electrode intensities per sampling time, with the lines coloured by electrode, and the eye status annotated using dark grey blocks.

```{r plot_data}
melt <- reshape2::melt(eeg_data %>% dplyr::select(-split), id.vars=c("eyeDetection", "ds"), variable.name = "Electrode", value.name = "microvolts")


ggplot2::ggplot(melt, ggplot2::aes(x=ds, y=microvolts, color=Electrode)) + 
  ggplot2::geom_line() + 
  ggplot2::ylim(3500,5000) + 
  ggplot2::geom_vline(ggplot2::aes(xintercept=ds), data=dplyr::filter(melt, eyeDetection==1), alpha=0.005)
```

**2** Do you see any obvious patterns between eyes being open (dark grey blocks in the plot) and the EEG intensities?\
Yes, there may be some patterns.
When the eyes are open, brain signals appear calmer and more stable.
Conversely, when the eyes are closed, the signals become more active and volatile

**3** Similarly, based on the distribution of eye open/close state over time to anticipate any temporal correlation between these states?\
Yes, there appears to be temporal correlation.
The eye states don’t change randomly but occur in blocks or periods, indicating that the person kept their eyes closed or open for extended periods rather than rapidly switching states.
This creates temporal dependencies, where the current eye state is highly predictive of the next state.

Let's see if we can directly look at the distribution of EEG intensities and see how they related to eye status.

As there are a few extreme outliers in voltage we will use the `dplyr::filter` function to remove values outwith of 3750 to 50003.
The function uses the `%in%` operator to check if each value of microvolts is within that range.
The function also uses the `dplyr::mutate()` to change the type of the variable eyeDetection from numeric to a factor (R's categorical variable type).

```{r compare_distrib}
melt_train <- reshape2::melt(eeg_train, id.vars=c("eyeDetection", "ds"), variable.name = "Electrode", value.name = "microvolts")

# filter huge outliers in voltage
filt_melt_train <- dplyr::filter(melt_train, microvolts %in% (3750:5000)) %>% dplyr::mutate(eyeDetection=as.factor(eyeDetection))

ggplot2::ggplot(filt_melt_train, ggplot2::aes(y=Electrode, x=microvolts, fill=eyeDetection)) + ggplot2::geom_boxplot()
```

Plots are great but sometimes so it is also useful to directly look at the summary statistics and how they related to eye status.
We will do this by grouping the data based on eye status and electrode before calculating the statistics using the convenient `dplyr::summarise` function.

```{r compare_summary_stats}
filt_melt_train %>% dplyr::group_by(eyeDetection, Electrode) %>% 
    dplyr::summarise(mean = mean(microvolts), median=median(microvolts), sd=sd(microvolts)) %>% 
    dplyr::arrange(Electrode)
```

**4** Based on these analyses are any electrodes consistently more intense or varied when eyes are open?\
The boxplots and summary statistics show that electrode responses differ between eye states.
For example, occipital electrodes (O1 and O2) have higher standard deviations when eyes are closed, indicating increased alpha wave activity.
Frontal electrodes (AF3 and AF4) may have varying mean intensities across eye states.
Posterior electrodes generally show greater variation during closed eyes.

#### Time-Related Trends

As it looks like there may be a temporal pattern in the data we should investigate how it changes over time.

First we will do a statistical test for stationarity:

```{r convert_to_tseries}
apply(eeg_train, 2, tseries::adf.test)
```

**5** What is stationarity?\
Stationarity is a statistical property of a time series where mean, variance, and autocorrelation remain constant over time.
A stationary time series has a constant mean and variance, and autocovariance depends only on lag, not time.
Many statistical models assume stationarity.

**6** Why are we interested in stationarity?
What do the results of these tests tell us?
(ignoring the lack of multiple comparison correction...)\
We are interested in it because stationarity is important because: many time series models (like ARIMA) need it, non-stationary data can cause spurious correlations, and stationary data is easier to model and predict.\
The test results likely indicate that most EEG channels reject the null hypothesis of non-stationarity, as indicated by low p-values.
This suggests that the data is stationary.
However, it’s important to note that this is typical for EEG data over short time periods, and eye state changes could potentially introduce some non-stationarity.

Then we may want to visually explore patterns of autocorrelation (previous values predict future ones) and cross-correlation (correlation across channels over time) using `forecast::ggAcf` function.

The ACF plot displays the cross- and auto-correlation values for different lags (i.e., time delayed versions of each electrode's voltage timeseries) in the dataset.
It helps identify any significant correlations between channels and observations at different time points.
Positive autocorrelation indicates that the increase in voltage observed in a given time-interval leads to a proportionate increase in the lagged time interval as well.
Negative autocorrelation indicates the opposite!

```{r correlation}
forecast::ggAcf(eeg_train %>% dplyr::select(-ds))
```

**7** Do any fields show signs of strong autocorrelation (diagonal plots)?
Do any pairs of fields show signs of cross-correlation?
Provide examples.\
The plot reveals autocorrelations along the diagonal and potential cross-correlations in the off-diagonal tiles for various EEG channels and the eyeDetection variable.
Several EEG channels, such as FC5, T7, P7, O1, O2, P8, T8, and FC6, exhibit strong autocorrelation.
This is evident from the repeating bar patterns that rise above the blue significance threshold lines, indicating these signals have regular, repeating patterns over time.
The eyeDetection variable also shows strong autocorrelation, which is expected since eye states often follow a regular open-close pattern in experimental designs.\
In terms of cross-correlation, a few pairs of channels exhibit notable relationships.
For instance, the F7 and eyeDetection pair displays a visible pattern that crosses the confidence threshold, suggesting a possible correlation between brain activity at the F7 location and eye openness.
Similarly, channels O1 and O2, located in the occipital region responsible for visual processing, show signs of cross-correlation, which aligns with their anatomical proximity and functional similarity.
Additionally, there is evidence of cross-correlation between P8 and T8, both in the temporal-parietal area.
These patterns indicate that certain brain regions not only exhibit strong internal signal consistency over time but also interact with one another or with behavioural states like eye openness.

#### Frequency-Space

We can also explore the data in frequency space by using a Fast Fourier Transform.\
After the FFT we can summarise the distributions of frequencies by their density across the power spectrum.
This will let us see if there any obvious patterns related to eye status in the overall frequency distributions.

```{r fft_open}
eegkit::eegpsd(eeg_train %>% dplyr::filter(eyeDetection == 0) %>% dplyr::select(-eyeDetection, -ds), Fs = Fs, xlab="Eye Open")
```

```{r fft_closed}
eegkit::eegpsd(eeg_train %>% dplyr::filter(eyeDetection == 1) %>% dplyr::select(-eyeDetection, -ds), Fs = Fs, xlab="Eye Closed")
```

**8** Do you see any differences between the power spectral densities for the two eye states?
If so, describe them.\
Yes there are differences:\
1.
Alpha rhythm (8-12 Hz): Much stronger in the eye closed condition (red/orange bands) compared to eye open (blue/cyan)\
2.
Overall power: Eye closed shows higher power and more organized frequency bands, while eye open has more uniform, lower-amplitude activity\
3.
Pattern structure: Eye closed displays distinct, band-like patterns; eye open appears more diffuse and less structured\
This reflects the classic EEG finding that alpha rhythms are enhanced when eyes are closed and suppressed during visual processing when eyes are open.

#### Independent Component Analysis

We may also wish to explore whether there are multiple sources of neuronal activity being picked up by the sensors.\
This can be achieved using a process known as independent component analysis (ICA) which decorrelates the channels and identifies the primary sources of signal within the decorrelated matrix.

```{r ica, warning=FALSE}
ica <- eegkit::eegica(eeg_train %>% dplyr::select(-eyeDetection, -ds), nc=3, method='fast', type='time')
mix <- dplyr::as_tibble(ica$M)
mix$eyeDetection <- eeg_train$eyeDetection
mix$ds <- eeg_train$ds

mix_melt <- reshape2::melt(mix, id.vars=c("eyeDetection", "ds"), variable.name = "Independent Component", value.name = "M")


ggplot2::ggplot(mix_melt, ggplot2::aes(x=ds, y=M, color=`Independent Component`)) + 
  ggplot2::geom_line() + 
  ggplot2::geom_vline(ggplot2::aes(xintercept=ds), data=dplyr::filter(mix_melt, eyeDetection==1), alpha=0.005) +
  ggplot2::scale_y_log10()
```

**9** Does this suggest eye opening relates to an independent component of activity across the electrodes?\
Yes, this plot suggests that eye activity is related to an independent component.
The sharp downward spikes represent eye blinks or movements that ICA has successfully isolated into a separate component from brain signals.
This component captures the electrical artifacts from eye movements, while meaningful neural changes, such as alpha rhythm differences, from eye opening and closing would appear in different, brain-related components.
In essence, ICA separates eye artifacts from the actual neural correlates of visual processing.

### Eye Opening Prediction

Now that we've explored the data let's use a simple model to see how well we can predict eye status from the EEGs:

```{r xgboost}
# Convert the training and validation datasets to matrices
eeg_train_matrix <- as.matrix(dplyr::select(eeg_train, -eyeDetection, -ds))
eeg_train_labels <- as.numeric(eeg_train$eyeDetection) -1

eeg_validate_matrix <- as.matrix(dplyr::select(eeg_validate, -eyeDetection, -ds))
eeg_validate_labels <- as.numeric(eeg_validate$eyeDetection) -1

# Build the xgboost model
model <- xgboost(data = eeg_train_matrix, 
                 label = eeg_train_labels,
                 nrounds = 100,
                 max_depth = 4,
                 eta = 0.1,
                 objective = "binary:logistic")

print(model)
```

**10** Using the `caret` library (or any other library/model type you want such as a naive Bayes) fit another model to predict eye opening.

```{r model2}
# Prepare data for caret
train_data <- eeg_train %>% dplyr::select(-ds)
validate_data <- eeg_validate %>% dplyr::select(-ds)

# Set up training control
ctrl <- trainControl(method = "none")  # No cross-validation since we have separate validation set

# Train Random Forest model
rf_model <- train(eyeDetection ~ ., 
                  data = train_data,
                  method = "rf",
                  trControl = ctrl,
                  ntree = 100)

print(rf_model)

# Make predictions and evaluate both models on validation set
xgb_pred_val <- predict(model, eeg_validate_matrix)
xgb_pred_val_class <- ifelse(xgb_pred_val > 0.5, 1, 0)

rf_pred_val <- predict(rf_model, validate_data)
rf_pred_val_numeric <- as.numeric(rf_pred_val) - 1

# Calculate validation accuracy
xgb_val_acc <- mean(xgb_pred_val_class == eeg_validate_labels)
rf_val_acc <- mean(rf_pred_val_numeric == eeg_validate_labels)

cat("XGBoost validation accuracy:", xgb_val_acc, "\n")
cat("Random Forest validation accuracy:", rf_val_acc, "\n")
```

**11** Using the best performing of the two models (on the validation dataset) calculate and report the test performance (filling in the code below):

```{r test}
# Prepare test data
eeg_test_matrix <- as.matrix(dplyr::select(eeg_test, -eyeDetection, -ds))
eeg_test_labels <- as.numeric(eeg_test$eyeDetection) - 1
test_data <- eeg_test %>% dplyr::select(-ds)

# Determine which model performed better on validation
if(xgb_val_acc > rf_val_acc) {
  # Use XGBoost
  test_pred <- predict(model, eeg_test_matrix)
  test_pred_class <- ifelse(test_pred > 0.5, 1, 0)
  best_model <- "XGBoost"
} else {
  # Use Random Forest
  test_pred_rf <- predict(rf_model, test_data)
  test_pred_class <- as.numeric(test_pred_rf) - 1
  best_model <- "Random Forest"
}

# Calculate test performance
test_accuracy <- mean(test_pred_class == eeg_test_labels)
confusion_matrix <- table(Predicted = test_pred_class, Actual = eeg_test_labels)

cat("Best model:", best_model, "\n")
cat("Test accuracy:", test_accuracy, "\n")
print("Confusion Matrix:")
print(confusion_matrix)

# Calculate additional metrics
precision <- confusion_matrix[2,2] / sum(confusion_matrix[2,])
recall <- confusion_matrix[2,2] / sum(confusion_matrix[,2])
f1_score <- 2 * (precision * recall) / (precision + recall)

cat("Precision:", precision, "\n")
cat("Recall:", recall, "\n")
cat("F1-score:", f1_score, "\n")
```

**12** Describe 2 possible alternative modeling approaches for prediction of eye opening from EEGs we discussed in the lecture but haven't explored in this notebook.\
Gaussian Processes can predict brain electrical signal patterns over time.
They handle noisy EEG data and provide confidence measures.
They consider electrode proximity and can detect mental states like eye opening and closing.
They’re powerful for small to medium EEG datasets with uncertainty.\
Deep Learning and Neural Networks, particularly Convolutional Neural Networks (CNNs) and Recurrent Neural Networks (RNNs/LSTMs), can capture spatial and temporal patterns in EEG data, respectively.
These approaches automatically learn relevant features from raw EEG signals.

**13** What are 2 R libraries you could use to implement these approaches?
(note: you don't actually have to implement them though!)\
Gaussian Processes — kernlab\
Deep Learning Neural Networks — keras

## Optional

**14** (Optional) As this is the last practical of the course - let me know how you would change future offerings of this course.
This will not impact your marks!

-   What worked and didn’t work for you (e.g., in terms of the practicals, tutorials, and lectures)?\
    I particularly enjoyed the blend of a variety of topics each week and the chance to apply them in the tutorials.

-   Was learning how to run the practicals on your own machines instead of a clean server that will disappear after the course worth the technical challenges?\
    Yes it was!

-   What would you add or remove from the course?\
    I would opt for Python over R, because it's more beginner friendly and also used more widely in the present industry compared to R.

-   What was the main thing you will take away from this course?\
    Planning for a research project is much more than having a big idea, but there are lots of steps and factors to account for.
    Many things like budget, conflict of interest, previous research, constraints and limitations need to be weighed.
